MS&E 226: Project Part 1 Code

### 2024-11-19 MS&E 226_Project Part 1_Code.Rmd
comments added

@@

# load data
df <- read.csv("/Users/sydneyloh/OneDrive/Documents/Stanford/MS&E 226/Term Project/electricity.csv")

@@

@@

```{r Correlations, fig.show='hide', results='hide'}

@@

ggplot(sorted_correlations_revenue, aes(x = reorder(Variable, Correlation_with_revenue), 
                                             y = Correlation_with_revenue)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +  # Flip coordinates for better readability
  labs(title = "Correlation with Revenue.Total", 
       x = "Variables", 
       y = "Correlation Coefficient") +
  theme_minimal()

# Calculate correlations with binary outcome variable
correlations_with_uses_consumed <- correlations["Uses.Consumed.Binary",]

@@

```{r 5f subgroups in data, fig.show='hide', results='hide'}

@@

View(state_means)
View(type_means)
View(region_means)
variables_to_plot <- names(state_means)[-1]  # Exclude Utility.State
plot_list <- list()

# Loop through each variable and create a plot
for (var in variables_to_plot) {
  p <- ggplot(state_means, aes_string(x = "Utility.State", y = var)) +
    geom_bar(stat = "identity", fill = "green") +
    coord_flip() +
    labs(title = paste("Mean", var, "by Utility State"),
         x = "Utility State",
         y = paste("Mean", var)) +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 6, angle = 0),
          axis.title.y = element_text(size = 10),
          axis.title.x = element_text(size = 10),
          plot.title = element_text(size = 12, face = "bold"))

@@

for (plot in plot_list) {
  print(plot)
}
# Create plots for utility type means
variables_to_plot_type <- names(type_means)[-1]  # Exclude Utility.Type
plot_list_type <- list()

@@

for (plot in plot_list_type) {
  print(plot)
}
# Define the list of variables to plot, excluding the grouping column (Utility.Region)
variables_to_plot <- names(region_means)[-1]

@@

# Print each plot for region_means
for (plot in plot_list) {
  print(plot)
}

@@

```{r regression}
# set seed
set.seed(123)

@@

columns_to_exclude <- c("Utility.Name", "Utility.Number", "Utility.State", 
                         "Revenue.Retail", "Revenue.Delivery", 
                         "Revenue.Resale", "Revenue.Adjustments", 
                         "Revenue.Transmission", "Revenue.Other", "Uses.Consumed") #Revenue subparts are linearly dependent on total revenue

# Create a new data frame without the excluded columns
# df_pred <- df[, !names(df) %in% columns_to_exclude]

@@

# Histogram of log-transformed Revenue.Total (after scaling)
ggplot(df_pred, aes(x = Log_Revenue_Total)) + 
  geom_histogram(binwidth = 0.1, fill = "lightgreen", color = "black", alpha = 0.7) + 
  labs(title = "Histogram of Log-transformed Revenue Total", 
       x = "Log(Revenue Total)", 
       y = "Frequency") +
  theme_minimal()
### Baseline Model
# Fit OLS Model (Linear Regression) without cross-validation
ols_model <- lm(Log_Revenue_Total ~ ., 
                data = df_pred)  # Fit the model on the entire dataset

@@

# Calculate RMSE (Root Mean Squared Error)
rmse <- sqrt(mean(residuals^2))
cat(paste("OLS Model RMSE:", rmse, "\n"))

### Cross Validation
# Set up cross-validation control with 10 folds
train_control <- trainControl(method = "cv", 
                              number = 10, 
                              summaryFunction = defaultSummary,
                              search = "grid",
                              savePredictions = "all")
### Ridge Regression
ridge_model <- train(Log_Revenue_Total ~ ., 
                     data = df_pred, 
                     method = "glmnet",  # glmnet is used for regularized regression
                     trControl = train_control,
                     metric = "RMSE", 
                     preProcess = "scale",  # Standardize the data
                     tuneGrid = expand.grid(alpha = 0, lambda = seq(0, 1, length = 100))) # alpha=0 for Ridge

@@

final_lambda_ridge <- ridge_model$bestTune$lambda
cat(paste("Final lambda value for Ridge regression:", final_lambda_ridge, "\n"))
### Lasso Regression
# Fit Lasso regression using glmnet with tuning of lambda and data standardization
lasso_model <- train(Log_Revenue_Total ~ ., 
                     data = df_pred, 
                     method = "glmnet",  # glmnet for regularized regression
                     trControl = train_control,
                     metric = "RMSE",  # RMSE for model evaluation
                     preProcess = "scale",  # Standardize the data
                     tuneGrid = expand.grid(alpha = 1, lambda = seq(0, 1, length = 100)))  # Lasso: alpha = 1

@@

# Extract and print the best lambda value for Lasso regression
final_lambda_lasso <- lasso_model$bestTune$lambda
cat(paste("Final lambda value for Lasso regression:", final_lambda_lasso, "\n"))
### Random Forest
# Fit a Random Forest model
rf_model <- train(Log_Revenue_Total ~ ., 
                  data = df_pred, 
                  method = "rf",  # Random Forest regression
                  trControl = train_control,
                  metric = "RMSE",
                  tuneGrid = expand.grid(mtry = sqrt(ncol(df_pred)))) # Hyperparameter tuning for mtry

@@

rf_model_rmse <- min(rf_model$results$RMSE)
cat(paste("Random Forest RMSE:", rf_model_rmse, "\n"))

@@

@@

@@

@@

@@

>>>>

# 2: Data Pre-processing

@@

# load data
df <- read.csv("electricity.csv")

@@

# 3: Hold Out Set

@@

# 5: Data Exploration

## Correlations

```{r Correlations with Revenue.Total, fig.show='hide', results='hide'}

@@

ggplot(sorted_correlations_revenue, aes(x = reorder(Variable, Correlation_with_revenue), 
                                             y = Correlation_with_revenue)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +  # Flip coordinates for better readability
  labs(title = "Correlation with Revenue.Total", 
       x = "Variables", 
       y = "Correlation Coefficient") +
  theme_minimal()
```

```{r Correlations with Uses.Consumed.Binary, fig.show='hide', results='hide'}
# Calculate correlations with binary outcome variable
correlations_with_uses_consumed <- correlations["Uses.Consumed.Binary",]

@@

## Subgroups in data

```{r means across categories, fig.show='hide', results='hide'}

@@

View(state_means)
View(type_means)
View(region_means)
variables_to_plot <- names(state_means)[-1]  # Exclude Utility.State
plot_list <- list()
```

```{r Utility.State plots, fig.show='hide', results='hide'}
# Loop through each variable and create a plot
for (var in variables_to_plot) {
  p <- ggplot(state_means, aes_string(x = "Utility.State", y = var)) +
    geom_bar(stat = "identity", fill = "green") +
    coord_flip() +
    labs(title = paste("Mean", var, "by Utility State"),
         x = "Utility State",
         y = paste("Mean", var)) +
    theme_minimal() +
    theme(axis.text.y = element_text(size = 6, angle = 0),
          axis.title.y = element_text(size = 10),
          axis.title.x = element_text(size = 10),
          plot.title = element_text(size = 12, face = "bold"))

@@

for (plot in plot_list) {
  print(plot)
}

```{r Utility.Type plots, fig.show='hide', results='hide'}
# Create plots for utility type means
variables_to_plot_type <- names(type_means)[-1]  # Exclude Utility.Type
plot_list_type <- list()

@@

for (plot in plot_list_type) {
  print(plot)
}

```{r Utility.Region Plots, fig.show='hide', results='hide'}
# Define the list of variables to plot, excluding the grouping column (Utility.Region)
variables_to_plot <- names(region_means)[-1]

@@

# Print each plot for region_means
for (plot in plot_list) {
  print(plot)
}
```

@@

## Data Transformations

```{r data transformations}
# set seed
set.seed(123)

@@

@@

# Histogram of log-transformed Revenue.Total (after scaling)
ggplot(df_pred, aes(x = Log_Revenue_Total)) + 
  geom_histogram(binwidth = 0.1, fill = "lightgreen", color = "black", alpha = 0.7) + 
  labs(title = "Histogram of Log-transformed Revenue Total", 
       x = "Log(Revenue Total)", 
       y = "Frequency") +
  theme_minimal()

## Baseline Model

```{r Baseline Model}
### Baseline Model
# Fit OLS Model (Linear Regression) without cross-validation
ols_model <- lm(Log_Revenue_Total ~ ., 
                data = df_pred)  # Fit the model on the entire dataset

@@

# Calculate RMSE (Root Mean Squared Error)
rmse <- sqrt(mean(residuals^2))
cat(paste("OLS Model RMSE:", rmse, "\n"))
```

## Cross Validation setup

```{r cross validation}
### Cross Validation
# Set up cross-validation control with 10 folds
train_control <- trainControl(method = "cv", 
                              number = 10, 
                              summaryFunction = defaultSummary,
                              search = "grid",
                              savePredictions = "all")
```

## Model 1: Ridge Regression

```{r Model 1: Ridge Regression}
### Ridge Regression
ridge_model <- train(Log_Revenue_Total ~ ., 
                     data = df_pred, 
                     method = "glmnet",  # glmnet is used for regularized regression
                     trControl = train_control,
                     metric = "RMSE", 
                     preProcess = "scale",  # Standardize the data
                     tuneGrid = expand.grid(alpha = 0, lambda = seq(0, 1, length = 100))) # alpha=0 for Ridge

@@

final_lambda_ridge <- ridge_model$bestTune$lambda
cat(paste("Final lambda value for Ridge regression:", final_lambda_ridge, "\n"))

## Model 2: Lasso Regression

```{r Model 2: Lasso Regression}
### Lasso Regression
# Fit Lasso regression using glmnet with tuning of lambda and data standardization
lasso_model <- train(Log_Revenue_Total ~ ., 
                     data = df_pred, 
                     method = "glmnet",  # glmnet for regularized regression
                     trControl = train_control,
                     metric = "RMSE",  # RMSE for model evaluation
                     preProcess = "scale",  # Standardize the data
                     tuneGrid = expand.grid(alpha = 1, lambda = seq(0, 1, length = 100)))  # Lasso: alpha = 1

@@

# Extract and print the best lambda value for Lasso regression
final_lambda_lasso <- lasso_model$bestTune$lambda
cat(paste("Final lambda value for Lasso regression:", final_lambda_lasso, "\n"))

## Model 3: Random Forest

```{r Model 3: Random Forest}
### Random Forest
# Fit a Random Forest model
rf_model <- train(Log_Revenue_Total ~ ., 
                  data = df_pred, 
                  method = "rf",  # Random Forest regression
                  trControl = train_control,
                  metric = "RMSE",
                  tuneGrid = expand.grid(mtry = sqrt(ncol(df_pred)))) # Hyperparameter tuning for mtry

@@

rf_model_rmse <- min(rf_model$results$RMSE)
cat(paste("Random Forest RMSE:", rf_model_rmse, "\n"))
```

@@

## Data Cleaning

@@

## Classifier 1: K-NN

@@

## Classifier 2: Logistic Regression with Feature Selection

@@

## Classifier 3: Decision Tree

@@

## Comparing Classifiers
